Using Energy Smart Meter Data and Socio-Demographic Indicators to Understand Consumption in Relation to Deprivation Levels

### 2018-06-03 paper/Paper_1_v02.Rmd
images

The UK continues to see domestic energy usage levels increase as an overall percentage of the countries consumption. In 2016, it accounted for 29% of overall usage - a decrease from previous years 35% in 2013 - but higher than the EU average which stood at 25.45% [@Eurostat]. These fluctuations are part of a larger upward trend overall, which has seen domestic energy consumption increase markedly from the 1970s by 12%, with short term fluctuations being accounted for by the effect of mean air temperature year on year [@Strategy].

Previous research has shown that a series of socio-demographic attributes influence the extent to which energy usage can be understood in the context of derpivation levels, including income, housing tenure and number of appliances. Indeed other research also covers factors such as dwelling characteristics and psychological and behavioural factors behind consumption patterns, which whilst important, are beyond the scope of this paper. Spatial variations have also previously been examined, but largely in reference to weather and temperature impacting upon consumption.

Primarily in response to concerns over reaching the targets laid out in the 2008 climate change act, central policy measures have been introduced to support and encourage reductions in energy usage, particularly in the domestic setting. Smart Meters are high on the UK government's energy efficiency agenda in order to align with those proposed targets. The UK government set goals of a 29% decrease in the domestic category [@HMGovernment2009] and for the transition to smart metering to be complete for both gas and electriciy by 2020 [@haben2016analysis]. With a top - down outlook, by upgrading the UK infrastructure to a 'Smart Grid' there is the potential for "solving a lot of our energy problems" [@smartgridcom; @Stern2011].Smart meters enable the smart grid by providing real-time data on the consumption of energy [@smartgridcom; @guerreiro2015making], therfore it will improve reliability, better match supply and demand and allow more efficient planning for both short and long term future; by being more secure, planning ahead for unexpected power outages and informing the number of power stations the UK is likely to need in the future.

Given the establishment of these targets, there is an operational setting in which Domestic Energy Providers (DEPs) are under increasing pressure to ensure their customers are on the most efficient and affordable tariffs, driving a demand for smart meter installations and bringing an end to pre-paid meters and estimated bills, which are often far more expensive than actual bills, but are also the most likely tariffs to be held by those in or near fuel poverty. These personalised tariffs will enable the DEPs to offer access to cheap energy when demand is at its lowest.

@@

@@

FIGURE 1 - previously FIG 4.

@@

FIGURE 2 - previously FIG 5

@@

FIGURE 3 - previously FIG 1

There is also a clear AM peak between 07:30 and 08:00, confirming that energy usage increases as people begin their daily tasks, but dips off again as people go out to work. The lowest energy usage occurs between the hours of 01:00 and 04:30, but does not go as low as zero because of the ‘standby’ rate of energy usage, where only those appliances which are not switched off such as the fridge freezer or central heating are consuming energy (Wyatt 2013). The fact that energy usage does not dip this low at any other time of the day indicates that in some cases energy is consumed throughout the day because people are at home through illness, unemployement, caring duties or shift work.

@@

FIGURES 4 and 5 - previously figures 2 and 3

@@

It is also true to say that the UKs housing stock is currently much older than that in the rest of Europe. This goes some way to explaining the relative higher consumption but also at an individual level, post-war terrace houses are still extremely common, but are poorly insulated and in need of updating [@DECC2017]. Those families living in large detached and semi-detached houses are also more likely to be owner occupiers, and the DECC stated in their latest report that it is the disconnect between tenants and landlords that stands in the way of privately rented homes becoming more energy efficient, because it is unclear who should pay for and benefit from the efficiency measures. Private renters are the most likely to be in the deepest fuel poverty.

Accommodation type is also indicative

The model ran attempts to explain the energy consumption by levels of deprivation.

The deprivation model ran here included the proportion of accomodation types in each PCS; detached, semi-detached, terrace, flat or shared accomodation. Due to the size of two of the groups, flats and shared accommodation were combined. As the largest category overall, semi-detached was used as the reference category. As can be seen in the OAC group pen profiles, deprivation and housing type are closely linke so it was important to control for it to avoid the Townsend score returning a significant result as it is being used as a proxy for type of accomodation.

>>>>

The UK continues to see domestic energy usage levels increase as an percentage of the countries overall consumption. In 2016, it accounted for 29% of overall usage - a decrease from previous years (35% in 2013) but higher than the EU average which stood at 25.45% [@Eurostat]. These fluctuations are part of a larger overall upward trend , which has seen domestic energy consumption increase markedly from the 1970s by 12%, with short term fluctuations being accounted for by the effect of mean air temperature year on year [@Strategy].

Previous research has shown that a series of socio-demographic attributes influence the extent to which energy usage can be understood in the context of derpivation levels, including income, housing tenure and number of appliances. Indeed other research also covers factors such as dwelling characteristics, psychological factors and behavioural factors behind consumption patterns which whilst important, are beyond the scope of this paper. Spatial variations have also previously been examined, but largely in reference to weather and temperature impacting upon consumption.

Primarily in response to concerns over reaching the targets laid out in the 2008 climate change act, central policy measures have been introduced to support and encourage reductions in energy usage, particularly in the domestic setting. Smart Meters are high on the UK government's energy efficiency agenda in order to align with those proposed targets. The UK government set goals of a 29% decrease in the domestic category [@HMGovernment2009] and for the transition to smart metering to be complete for both gas and electriciy by 2020 [@haben2016analysis]. From a top down approach, by upgrading the UK infrastructure to a 'Smart Grid' there is the potential for "solving a lot of our energy problems" [@smartgridcom; @Stern2011].Smart meters enable the smart grid by providing real-time data on the consumption of energy [@smartgridcom; @guerreiro2015making], therfore it will improve reliability, better match supply and demand and allow more efficient planning for both short and long term future; by being more secure, planning ahead for unexpected power outages and informing the number of power stations the UK is likely to need in the future.

Given the establishment of these targets, there is an operational setting in which Domestic Energy Providers (DEPs) are under increasing pressure to ensure their customers are on the most efficient and affordable tariffs, driving a demand for smart meter installations and bringing an end to pre-paid meters and estimated bills, which are often far more expensive than actual bills, but are also the most likely tariffs to be held by those in or near fuel poverty [@DECC2017]. These personalised tariffs will enable the DEPs to offer access to cheap energy when demand is at its lowest.

@@

The 2008 Climate Change Act in the UK requires a 34% reduction in 1990 greenhouse gas emissions by 2020 and an 80% reduction by 2050. Dwellings are an important target area for reductions and the UK government set the goal of a 29% decrease in this area by 2020 [@HMGovernment2009]. As such, the installation of smart meter technologies that enable the collection of digital records of gas and electricity consumption are viewed as integral to encouraging greater efficiency through the measurement and monitoring of consumer behaviour alongside more efficiently producing and storing resources to meet the demand [@UKCCC2010; @guerreiro2015making; @smartgridcom].

@@

```{r SM_PROP, echo=FALSE, fig.cap="Smart Meter Penetration in the UK", out.width = '100%'}
knitr::include_graphics('SM_PROP.jpg')
```

@@

```{r LISA, echo=FALSE, fig.cap="Local Morans I", out.width = '100%'}
knitr::include_graphics("Local_morans_I.jpg")
```

@@

```{r hourly, echo=FALSE, fig.cap="Hourly rates of Total Energy Consumption", out.width = '100%'}
knitr::include_graphics("Hourly.jpg")
```
There is also a clear AM peak between 07:30 and 08:00, confirming that energy usage increases as people begin their daily tasks, but dips off again as people go out to work. The lowest energy usage occurs between the hours of 01:00 and 04:30, but does not go as low as zero because of the ‘standby’ rate of energy usage, where only those appliances which are not switched off such as the fridge freezer or central heating are consuming energy (Wyatt 2013). The fact that energy usage does not dip this low at any other time of the day indicates that in some cases energy is consumed throughout the day because people are at home through illness, unemployement, caring duties or shift work.

@@

```{r daily, echo=FALSE, fig.cap="Daily rates of Total Energy Consumption", out.width = '100%'}
knitr::include_graphics("Daily.jpg")
```

```{r seasonal, echo=FALSE, fig.cap="Seasonal rates of Total Energy Consumption", out.width = '100%'}
knitr::include_graphics("Seasonal.jpg")
```

@@

It is also true to say that the UKs housing stock is currently much older than that in the rest of Europe. This goes some way to explaining the relative higher consumption but also at an individual level, post-war terrace houses are still extremely common, but are poorly insulated and in need of updating [@DECC2017]. Those families living in large detached and semi-detached houses are also more likely to be owner occupiers, and the DECC stated in their latest report that it is the disconnect between tenants and landlords that stands in the way of privately rented homes becoming more energy efficient, because it is unclear who should pay for and benefit from the efficiency measures. Private renters are the most likely to be in the deepest fuel poverty [@DECC2017].

The model attempts to explain the energy consumption by levels of deprivation. It included the proportion of accomodation types in each PCS; detached, semi-detached, terrace, flat or shared accomodation. Due to the size of two of the groups, flats and shared accommodation were combined into one to make them more representative. As the largest category overall, semi-detached was used as the control category. As can be seen in the OAC group pen profiles, deprivation and housing type are closely linked so it was important to control for it to avoid the Townsend score returning a significant result as it is being used as a proxy for type of accomodation.
